---
title: "Complexity Explorer - Introduction to Differential Equations - Unit 2 Homework"
author: "BGS"
date: "25 April 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Questions and Answers to Q3 (i) - (iii)
These are my answers to the Unit 2 homework, namely Q3 . I've made
use of some R packages in doing this. The whole course can be found at the link
below:

<https://www.complexityexplorer.org/tutorials/31-introduction-to-differential-equations/segments/incomplete?summary>

First of, I'm going to load some packages and set the working directory to set the
stage for answering these questions.
```{r}
  library(deSolve)
  library(ggplot2)
  library(reshape2)
  library(magrittr)
```
Set a working directory (i.e. where you want to store this or other files.
Usually has the form 'C:/Users/YOU/SOME_FOLDER' on Windows machines

  setwd("C:/Users/YOU/SOME_FOLDER") 

#### Q3. Write a program that implements Euler's method for the main example of this unit: 

dT/dt = 0.2 * (20 - T)

#### Some things to try out or experiment with:
#### (i) Have your program produce a plot of your Euler solution.

O.K so I cheat a little here (already!), by making use of the 'deSolve' package.
I'm going to break down the solution to this problem in the standard way I would
construct a file were I writing it as an *.R script
```{r}
################################################################################
# PARAMETERS
################################################################################

  parameters <- c(x <- 0.2,
                  y <- 20)

################################################################################
# STATE VARIABLES - ENSURE assignment operator is '=' NOT '<-' within 'state'
################################################################################

  state <- c(T = 10)

################################################################################
# MODEL
################################################################################

  temp <- function (t, state, parameters){
    with(as.list(c(state, parameters)),{ #this line converts the 'state' and 
      #'parameter' vectors to lists
      #rate of change
      dT <- x * (y - T)
      # return the rates of change
      return(list(dT))
    })  #end with list
  }

  # Q3(i)
  # Define times over which simulation is run
  # TIME OF SIMULATION
  times1 <- seq(0,40, by = 1)
  
################################################################################
# OUTPUT
################################################################################

  # Q3(i) - 'out' is the dataframe containing the results for Q3(i)
  out <- ode(state, times1, temp, parameters, method = rkMethod("euler"))
  df.out <- as.data.frame(out)

################################################################################
# PLOTTING
################################################################################

  #Q3(i)
  temp.p <- ggplot(df.out, aes(time,T)) + 
    geom_point()
```
Here is the answer / plot to Q3(i)
```{r temp.p, echo=FALSE}
temp.p
```

#### (ii) Make plots of the Euler solutions for several different values of delta t.

```{r}
################################################################################
# MODEL
################################################################################

# Q3(ii)
  # Now define the vector of timesteps (delta t) to try:
  stepvec <- c(0.1,1.0,10)
# and define a list to hold the results:
  res <- vector(length(stepvec),mode="list")

  # take each value of 'x' and create a sequence dependent on 'x'
  for(k in seq_along(stepvec)){
    res[[k]] <- ode(y=state, 
                    times=(seq(0,100,by=stepvec[k])),
                    func=temp,
                    parms=parameters)
  }
  
################################################################################
# OUTPUT
################################################################################

  # Q3(ii) - 'out2' is the dataframe containing the results for Q3(ii)
  names(res) <- stepvec  ## to get stepvec values incorporated in results
  out2 <- dplyr::bind_rows(lapply(res,as.data.frame),.id="step")
  out2$step <- as.numeric(out2$step)
  
################################################################################
# PLOTTING
################################################################################

  #Q3(ii)
  temp.p2 <- ggplot(out2, aes(time,T,colour=factor(step))) + 
    geom_point() +
    scale_color_discrete()
```

Here's the result to Q3(ii) with the trajectories overlaid by one another using different timesteps / values of delta t

```{r temp.p2, echo=FALSE}
temp.p2
```
```{r}
  #Q3(ii)
  temp.p3 <- ggplot(out2, aes(time,T,colour=factor(step))) + 
    geom_point() +
    scale_color_discrete() +
    facet_wrap(~step)
```

Here's the result to Q3(ii) with the trajectories seperated on the different timesteps / values of delta t 

```{r temp.p3, echo=FALSE}
temp.p3
```

#### (iii) Compare the Euler solution with the exact solution

First, it's necessary to create a function to calculate the analytical solution. Then, for the purposes of comparison, we combine the analytical and numerical solutions to gauge their similarity or otherwise.

```{r}
################################################################################
# ANALYTIC RESULT 
################################################################################

  # time vector
  times2 <- seq(1,100,0.1)
```
```{r echo=T, results='hide'}
  # function
  temp.analy <- function(y){
    T = 20 - (5 * (exp(-0.2 * y)))
    print(T)
  }
```
```{r echo=T, results='hide'}
  # using sapply to apply 'temp.analy' over time vector 'y'
  temp.res <- sapply(times2, temp.analy)
```
```{r}
  # changes to ANALYTICAL dataset for combing with numerical
  temp.res2 <- cbind.data.frame(times2,temp.res)
  colnames(temp.res2) <- c("time","T")
  temp.res2["code"] <- "analytical"

  # changes to NUMERICAL dataset for combing with analytical
  out2.sub <- (out2[1:1001,])
  out2.sub <- out2.sub[,-1]
  out2.sub["code"] <- "numerical" 

  # combine ANALYTICAL dataset (by row) with NUMERICAL dataset
  comb.d <- rbind.data.frame(temp.res2,out2.sub)
```

Q3(iii) plotted results are shown below;

```{r}
################################################################################
# PLOTTING
################################################################################

  # Analytic result plot
  temp.res2.p <- ggplot(temp.res2, aes(times2,temp.res)) +
    geom_point()
```

The plot below shows the analytical solution

```{r}
temp.res2.p
```
```{r}
  # Analytic versus Numerical result comparison plot
  comb.p <- ggplot(comb.d, aes(time,T,colour=code)) +
    geom_point() +
    scale_color_discrete()
```

The plot below shows a comparison between the analytical and numerical solutions.

```{r}
comb.p
```

Maybe you would like to save the data? If so run the code below

```{r}
################################################################################
# SAVE DATA
################################################################################

  save.image("CE.IDE.unit2_v3.RData")

################################################################################

################################################################################
```